---
title: "Data Management Example"
author: ""
date: "Spring 2015"
output: html_document
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(descr)
```

Three categorical variables relevant to *health and wealth* are `incomeperperson,	lifeexpectancy,	employrate` from the `Gapminder` data set.  The `Gapminder` data set is loaded from the `PDS` package.  Using the `dplyr` package, a new data frame `HNW` containing only three variables, is created.


```{r}
library(PDS)
library(dplyr)
HNW <- gapminder %>% 
  select(incomeperperson,  lifeexpectancy,	employrate)
head(HNW)
```

Responses of `23` for question `incomeperperson` are recoded to `NA`. Responses of `22` for question `lifeexpectancy` are recoded to `NA`. Responses of `35` for question `employrate` are recoded to `NA`. 

```{r}
xtabs(~incomeperperson, data = HNW)
HNW$incomeperperson[HNW$incomeperperson == 23] <- NA
xtabs(~incomeperperson, data = HNW)
HNW$incomeperperson <- factor(HNW$incomeperperson, )[, drop = TRUE]
xtabs(~incomeperperson, data = HNW)
summary(HNW$incomeperperson)
freq(HNW$incomeperperson)
freq(HNW$incomeperperson, main = "National Income Per Person")
ggplot(data = HNW, aes(x = incomeperperson, fill = incomeperperson)) + 
  geom_bar() + 
  labs(title = "National Income Per Person", x = "") +
  theme_bw() +
  guides(fill = guide_legend(title = "National Income Per Person"))
```


Changing the $y$-axis to percent:

```{r}
ggplot(data = HNW, aes(x = incomeperperson, y = (..count..)/sum(..count..), fill = incomeperperson)) + 
  geom_bar() + 
  labs(title = "Did you ever seriously think about committing suicide?", x = "") +
  theme_bw() +
  guides(fill = guide_legend(title = "Have You\nConsidered\nSuicide?")) +
  labs(y = "percent")
```

Another approach to graph the categorical variables is to use the base function `barplot()`.

```{r}
barplot(xtabs(~incomeperperson, data = HNW), col = c("green", "red", "orange", "brown"))
barplot(prop.table(xtabs(~incomeperperson, data = HNW)), col = c("green", "red", "orange", "brown"))
```

Question `TimesAttemptedSuicide` asks the 821 respondents from `incomeperperson` "how many times did you attempt suicide?"  Responses of 7 are recoded to `NA`.

```{r}
xtabs(~TimesAttemptedSuicide, data = HNW)
HNW$TimesAttemptedSuicide[HNW$TimesAttemptedSuicide==7] <- NA
xtabs(~TimesAttemptedSuicide, data = HNW)
HNW$TimesAttemptedSuicide <- factor(HNW$TimesAttemptedSuicide)[, drop = TRUE]
xtabs(~TimesAttemptedSuicide, data = HNW)
HNW$TimesAttemptedSuicide <- factor(HNW$TimesAttemptedSuicide, 
                          labels = c("0 times", "1 time", "2 or 3 times", 
                                     "4 or 5 times", ">= 6 times", "refused", 
                                     "don't know"))
```



```{r}
summary(HNW$TimesAttemptedSuicide)
xtabs(~TimesAttemptedSuicide, data = HNW)
NAS <- sum(xtabs(~TimesAttemptedSuicide, data = HNW)[2:5])  # number that attempted suicide
NAS
freq(HNW$TimesAttemptedSuicide, main = "How many times did you actually attempt suicide?")
ggplot(data = HNW, aes(x = TimesAttemptedSuicide)) + 
  geom_bar(fill = "maroon") + 
  labs(title = "How many times did you actually attempt suicide?", x= "") +
  theme_bw() 
ggplot(data = na.omit(HNW[ , "TimesAttemptedSuicide", drop = FALSE]), aes(x = TimesAttemptedSuicide)) + 
  geom_bar(fill = "plum") + 
  labs(title = "How many times did you actually attempt suicide?", x = "") +
  theme_bw() 
ggplot(data = na.omit(HNW[ , "TimesAttemptedSuicide", drop = FALSE]), 
       aes(x = TimesAttemptedSuicide, fill = TimesAttemptedSuicide)) + 
  geom_bar() + 
  labs(title = "How many times did you actually attempt suicide?", x = "") +
  theme_bw() +
  guides(fill = guide_legend(title = "Number of Times\nSuicide Attempted"))
```

Question `TreatedByDoctor` asks of the `r NAS` that attempted suicide "Did you require treatment by either a nurse or doctor?"  A response of 7 should be recoded to `NA`.

```{r}
HNW$TreatedByDoctor[HNW$TreatedByDoctor == 7] <- NA
HNW$TreatedByDoctor <- factor(HNW$TreatedByDoctor, labels = c("No", "Yes"))[, drop = TRUE]
xtabs(~TreatedByDoctor, data = HNW)
freq(HNW$TreatedByDoctor, main = "Did you require treatment by either a nurse or doctor?")
ggplot(data = na.omit(HNW[ , "TreatedByDoctor", drop = FALSE]), aes(x = TreatedByDoctor, fill = TreatedByDoctor)) + 
  geom_bar() + 
  labs(title = "Did you require treatment by either a nurse or doctor?", x = "") +
  theme_bw() +
  guides(fill = guide_legend(title = "Treated by Doctor\n or Nurse"))
```